---
title: "ST 558 Project 1"
author: "Jessica Speer"
date: "June 11, 2020"
output: 
  rmarkdown::github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("httr")
library("jsonlite")
library("tidyverse")
library("tidyr")
library("dplyr")
library("stringi")
library("ggplot2")
library("plyr")
```


# JSON overview

JSON, short for JavaScript Object Notation, is a text format that is commonly used when sending or retrieving data across servers and/or APIs. There are several reasons why JSON is commonly used. For one, JSON is language-indepdent which reduces many issues that can come from using different languages across platforms. JSON is also easily readable by both humans and machines, and it parses cleanly, which is very beneficial when pulling data from sources that may have various formatting and structure. There are plenty of R packages that can assist with reading JSON into R, which are explored in the next section. 

References:

https://www.json.org/json-en.html

https://en.wikipedia.org/wiki/JSON

https://www.w3schools.com/js/js_json_intro.asp


# Reading JSON Data into R

There are three major packages that can be used to read JSON data into R: `rjson`, `RJSONIO`, and `jsonlite`. The packages are similar, especially if using for simple tasks. `RJSONIO` was developed as an alternative to `rjson` due to limitations with speed and bandwidth, however, `rjason` has been improved over time. `jsonlite` started as a fork of `RJSONIO` but the manual (linked below) notes that recent versions have been completely rewritten.

For this project, I chose `jsonlite` because I have had experience with it before, and because they specifically mention powerful capability to interact with a web API.

References:

https://cran.r-project.org/web/packages/rjson/rjson.pdf

https://cran.r-project.org/web/packages/RJSONIO/index.html

https://cran.r-project.org/web/packages/jsonlite/jsonlite.pdf

# Function to Return Parsed Data from NHL API

The function that is generated below enables the user to pull several data sets from the NHL API.

**Usage:**

`getNHL(dataset, franchiseID)`

**Arguments:**

`dataset`: Enter `"franchise"`, `"teamtotals"`, `"seasonrecs"`, `"goalierecs"`, or `"skaterrecs"`

`franchiseID`: Valid Franchise ID required for `"seasonrecs"`, `"goalierecs"`, and `"skaterrecs"`

**Datasets:**

`"franchise"`: Returns id, firstSeasonId and lastSeasonId and name of every team in the history of the NHL

`"teamtotals"`: Returns total stats for every franchise (ex roadTies, roadWins, etc)

`"seasonrecs"`: Drill-down into season records for a specific franchise

`"goalierecs"`: Goalie records for the specified franchise

`"skaterrecs"`: (Skater records, same interaction as goalie endpoint

**Function generation:**

```{r, echo=TRUE, message=FALSE}
getNHL <- function(str1=NULL, id=NULL) {
  if (is.null(str1)){
    return("Please enter desired data and Franchise ID (if applicable)")
  } 
  else
  s1<-stri_compare(str1, "franchise")
  if (s1==0){
    f<-GET("https://records.nhl.com/site/api/franchise")
    f<-content(f, "text")
    f<-fromJSON(f,flatten=TRUE)
    f<-data.frame(f)
    f<-tbl_df(f)
    f <- f %>% select(data.id, data.firstSeasonId, data.lastSeasonId, data.teamCommonName)
    return(f)
  }
  else
  s2<-stri_compare(str1, "teamtotals")
  if (s2==0){
    f<-GET("https://records.nhl.com/site/api/franchise-team-totals")
    f<-content(f, "text")
    f<-fromJSON(f,flatten=TRUE)
    f<-data.frame(f)
    f<-tbl_df(f)
    return(f)
  }
  else
  s3<-stri_compare(str1, "seasonrecs")
  s4<-stri_compare(str1, "goalierecs")
  s5<-stri_compare(str1, "skaterrecs")
  if ((s3==0 | s4==0 | s5==0) & is.null(id)){
    return("Please enter a Franchise ID in the second argument.")
  }
  else
  `%notin%` <- Negate(`%in%`)
  if ((s3==0 | s4==0 | s5==0) & id %notin% 1:38){
    return("Franchise ID not recognized.")
  }  
  else
  if (s3==0 & id %in% 1:38){
    f<-GET(paste0("https://records.nhl.com/site/api/franchise-season-records?cayenneExp=franchiseId=", id))
    f<-content(f, "text")
    f<-fromJSON(f,flatten=TRUE)
    f<-data.frame(f)
    f<-tbl_df(f)
    return(f)
  }
  else
  if (s4==0 & id %in% 1:38){
    f<-GET(paste0("https://records.nhl.com/site/api/franchise-goalie-records?cayenneExp=franchiseId=", id))
    f<-content(f, "text")
    f<-fromJSON(f,flatten=TRUE)
    f<-data.frame(f)
    f<-tbl_df(f)
    return(f)
  }
  else
  if (s5==0 & id %in% 1:38){
    f<-GET(paste0("https://records.nhl.com/site/api/franchise-skater-records?cayenneExp=franchiseId=", id))
    f<-content(f, "text")
    f<-fromJSON(f,flatten=TRUE)
    f<-data.frame(f)
    f<-tbl_df(f)
    return(f)
  }  
}

```

## Using the `getNHL` Function

The function can be used to call several NHL data sets from the API. Examples are illustrated below.

**Call Franchise data, select variables**

```{r, echo=TRUE, message=FALSE}
getNHL("franchise") %>% select(data.firstSeasonId, data.teamCommonName) %>% head(n=10)
```

**Call Team Totals data, select variables**

```{r, echo=TRUE, message=FALSE}
getNHL("teamtotals") %>% select(data.teamName, data.homeWins) %>% head(n=10)
```

**Call Season Records data, select variables**

```{r, echo=TRUE, message=FALSE}
getNHL("seasonrecs", 38) %>% select(data.franchiseName, data.homeWinStreak)
```

**Call Goalie Records data, select variables**

```{r, echo=TRUE, message=FALSE}
getNHL("goalierecs", 38) %>% select(data.firstName, data.lastName, data.wins)
```

**Call Skater Records data, select variables**

```{r, echo=TRUE, message=FALSE}
getNHL("skaterrecs", 38) %>% select(data.firstName, data.lastName, data.points) %>% head(n=10)
```

# Explore the NHL Data

## Categorical and Numerical Summaries

We will start with the Team Totals dataset, which includes total stats for each franchise.  We will limit to game type "2" for simplicity.

```{r, echo=TRUE, message=FALSE}
t<-getNHL("teamtotals")
t<-t[t$data.gameTypeId==2,]
```

We will first explore the win ratios (wins/games played). We will need to generate and assign some variables. We will create a categorical variable to indicate whether a franchise had a below average win ratio or at/above average.

```{r, echo=TRUE, message=FALSE}
t$WinRatio<-t$data.wins/t$data.gamesPlayed
avgWins<-mean(t$WinRatio)
t$WinCat<-NA
t$WinCat[t$WinRatio < avgWins] <- "Below Avg Wins"
t$WinCat[t$WinRatio >= avgWins] <- "At/Above Avg Wins"
```

Let's take a look at the win ratio distributions overall and by win ratio category.

**Overall**

```{r, echo=TRUE, message=FALSE}
summary(t$WinRatio)
```

**Below Average Win Ratio**

```{r, echo=TRUE, message=FALSE}
summary(t$WinRatio[t$WinRatio < avgWins])
```

**At/Above average Win Ratio**

```{r, echo=TRUE, message=FALSE}
summary(t$WinRatio[t$WinRatio >= avgWins])
```

We can create a second categorical variable using a ratio of penalty minutes (penalty min/games played). This will allow us to demonstrate a contingency table.

```{r, echo=TRUE, message=FALSE}
t$PMinRatio<-t$data.penaltyMinutes/t$data.gamesPlayed
avgPMins<-mean(t$PMinRatio)
t$PMinCat<-NA
t$PMinCat[t$PMinRatio < avgPMins] <- "Below Avg Penalty Mins"
t$PMinCat[t$PMinRatio >= avgPMins] <- "At/Above Avg Penalty Mins"
t1<-table(t$WinCat, t$PMinCat)
knitr::kable(t1, row.names=TRUE, caption = "Avg Wins vs Avg Penality Min")
```

It appears that the at/above avg win ratio group shares a higher proportion with the above avg penalty minute ratio group, and the opposite is observed for the below avg win ratio group.

## Boxplots

Now let's move on to some graphics! We will start with boxplots. First, we will stack the data so that we can look at home wins and home losses side by side.

```{r, echo=TRUE, message=FALSE}
t$WinCat<-as.factor(t$WinCat)
homewins<-t %>% select(data.franchiseId, WinCat, homeWL=data.homeWins, data.gamesPlayed)
homewins<-data.frame(homewins)
homewins$WLcat<-"Home Wins"
homeloss<-t %>% select(data.franchiseId, WinCat, homeWL=data.homeLosses, data.gamesPlayed)
homeloss<-data.frame(homeloss)
homeloss$WLcat<-"Home Losses"
homedata<-rbind(homewins, homeloss)
homeWLratio<-homedata$homeWL/homedata$data.gamesPlayed
```

Generate boxplots of the home win/loss ratios, with the avg win level indicator variable as the facet wrap.

```{r, echo=TRUE, message=FALSE}
g <- ggplot(homedata, aes(x=WLcat, y=homeWLratio, fill=WLcat)) + geom_boxplot() + facet_wrap(homedata$WinCat)
g
```

It appears that the at/above avg win ratio group may have had more of a home field advantage than the below avg win ratio group, who's boxplot distributions are closer together between home wins and home losses.

## Column Graphs

We will now use the Goalie Records data set to demonstrate a column graph. For simplicty, we will limit to the Boston Bruins and to the goalies who have played 200 or more games. We will graph the total number of shut outs for each goalie.

```{r, echo=TRUE, message=FALSE}
rm(t)
f<-getNHL("goalierecs", 6)
f<-f[f$data.gamesPlayed>=200,]
g <- ggplot(f, aes(x=data.lastName, y=data.shutouts)) + geom_col(fill="darkblue")
g
```

The column graph is a nice way to get an overall sense of this data. We can see that of these goalies (who have played 200 or more games), Thompson has the most shut outs, followed by Rask and Brimsek.

## Scatter Plots

Finally, we will explore the Skater Records data to demonstrate scatter plots. We will plot assists vs. goals, split by active status (active=TRUE or FALSE). First, we will overlay using different colors, then we will split them out using `facet_wrap`.

```{r, echo=TRUE, message=FALSE}
s<-getNHL("skaterrecs", 6)
data.activePlayer<-as.factor(s$data.activePlayer)
g <- ggplot(s, aes(x=data.assists, y=data.goals, color=data.activePlayer)) + geom_point()
g
g <- ggplot(s, aes(x=data.assists, y=data.goals)) + geom_point(color="darkblue") + facet_wrap(s$data.activePlayer)
g
```

As expected, the number of assists appear to be correlated with number of goals. The patterns seem to follow similarly for both groups.
